Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells

### 2024-10-14 Manuscript.qmd
Published

---
title: "Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Ivan Briz i Godino
    affiliations:
      - name: Department of Evolutionary Biology, Ecology and Environmental Science, Biodiversity Research Institute (IRBio), University of Barcelona
        address: Avinguda Diagonal, 643
        city: Barcelona
        country: Spain
        postal-code: 08028 
abstract: |
  This study provides a reassessment of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST). While *Patella depressa* along the Spanish shoreline and *Patella caerulea* in the Mediterranean have repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species. *Patella vulgata*, *Nacella deaurata* have been studied using Mg/Ca ratios with mixed or contrary results. In this study, we present elemental maps of these two species as well as *Nacella magellanica* together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST. Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios. By reassessing these species and including modern and archaeological specimens (n=12) from three patelloid species (*P. vulgata*, *N. deaurata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change. As a result, these species can in the future serve to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O and clumped isotope analysis.
keywords: 
  - Sclerochronology
  - Patella
  - Nacella
  - LIBS
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: FALSE
    latex_engine: xelatex
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

@@

Limpet shells are commonly found within archaeological sites and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source [@Alvarez2011-rm, @Robson2023-oz]. They have been successfully studied in the context of coastal subsistence economies and site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scales [@Ortiz2015-mr], as well as archives of palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl]. Determining past sea surface temperature (SST) change recorded in patelloid limpet shells largely relies on the measurement of $\delta$^18^O values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) ratios [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure that potentially provides a more accurate SST estimate than $\delta$^18^O-values,as the latter proxy is affected by changes in the mixing of fresh- and saltwater. In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].

@@

Recent research particularly of *Patella* spp. in the Mediterranean and southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf]. However, there remains a lack of clarity for Atlantic limpet species, particularly since it was shown, that the Mg/Ca ratio in their shells are not reliable recorders of palaeotemperature [@Graniero2017-io]. Using modern reference material Graniero *et al*. [-@Granieo2017-io] tested multiple elemental ratios (Mg/Ca, Sr/Ca, Li/Ca, Li/Mg, and Sr/Li) acquired via LA-ICP-MS measurements with no ratios showing correlations with corresponding SST changes nor any sinusoidal patterns that would indicate a seasonal environmental control of elemental ratios. This study included two species also used in our study (*Patella vulgata*, *Nacella deaurata*), with one *P. vulgata* specimen (i.e. specimen ORK-LT5) being analysed by both. We also added specimens of the species *Nacella magellanica,* which — similar to the other two species — is also an established palaeothermometer using $\delta$^18^O-values [@Colonese2012-ct; @gutiérrez-zugasti2017; @Nicastro2020-ih].

In this study, we repeat and expand upon the analysis of Atlantic limpets to determine the reliability of Mg/Ca ratios as palaeotemperature proxies. To do this, we sampled a set of previously published and unpublished limpet specimens dating to modern and archaeological contexts using LIBS. LIBS allows us to carry out 2D imaging of entire shell sections, which facilitates navigation of the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio. By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest. Generally establishing the usefulness and reliability of Mg/Ca ratios as SST proxy in limpets provides a platform for future research and an important steppingstone to a better understanding of elemental ratios in other marine mollusc shells.

@@

| Context        | Study              | Species          | Location        | Sample ID  | Previous analyses                     |
|------------|------------|------------|------------|------------|------------|
| Modern         | [@Nicastro2020-ih] | *N. deaurata*    | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C |
|                |                    |                  | \- Tierra del   | ND-1016-4  | analysis                              |
|                |                    | *N. magellanica* | Fuego (AR)      | NM-1016-1  |                                       |
|                |                    |                  |                 | NM-1016-3  |                                       |
|                |                    |                  |                 |            |                                       |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | $\delta$^18^O and $\delta$^13^C |
|                |                    |                  | Westray (UK)    |            | analysis;                             |
|                |                    |                  |                 |            | Mg, Li, Sr, Ca.                       |
|                |                    |                  |                 |            |                                       |
|                |                    |                  |                 |            |                                       |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | $\delta$^18^O and $\delta$^13^C |
|                |                    |                  | Westray (UK)    | QG1-7189-2 | analysis                              |
|                |                    |                  |                 | QG2-1061-1 |                                       |
|                |                    |                  |                 | QG2-1064-1 |                                       |
|                |                    |                  |                 | QG2-7180-1 |                                       |
|                |                    |                  |                 | QG2-7180-2 |                                       |
|                |                    |                  |                 | QG1-7246-1 |                                       |

@@

Specimens were sourced from two main locations: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in the Orkney archipelago of Scotland (UK). Four modern specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016. The area receives around 570 mm of precipitation, but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu (practical salinity units) [@Nicastro2020-ih]. SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February (@fig-SSTs -- red line).

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for $\delta$^18^O and $\delta$$^13^C values, as well as via LA-ICP-MS [@Graniero2017-io]. Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho]. The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs -- blue line).

```{r SST_plot,fig.pos='H'}
#| fig-cap: Monthly sea surface temperature values averaged from 2007 to 2023. Values were sourced from @Good2020-nl.
#| fig-height: 3
#| fig-width: 6
#| label: fig-SSTs

@@

We also studied 7 archaeological specimens of *P. vulgata* from the Quoygrew site on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly (810 -- 1229 CE) [@Surge2012-ba].

@@

High-quality $\delta$^18^O values were acquired from previous publications on *P. vulgata* [@Surge2012-ba; @Graniero2017-io], *N. deaurata* and *N. magellanica* [@Nicastro2020-ih], in which detailed descriptions of the methodology for the stable isotope analyses can be found. In short, all shells were sectioned along the maximum growth axis to expose the internal growth structure. From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic layers of the shells. The exact location of the sampled carbonate powder is not available apart from the distance to the growth edge, where reported. Analytical precision of the isotope ratios was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.

@@

@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species (*Nacella* spp., and *Patella vulgata*, respectively). Due to the high-resolution sampling, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* spp. shells living 2–3 years and *P. vulgata* living regularly over 3 years. In those instances where longer lifespans resulted in slow growth rates below 1 mm/yr (i.e. specimen QG1-7246-1, @fig-Pat_iso b), the temporal-resolution is reduced from more than a dozen of samples per year, to only around 7. Together they show a range of lifespans, with some comparatively long lived for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to within-shell time-averaging because of slow growth rates. That said, the majority are under 5 years old yielding sufficiently fast growth to nearly capture seasonal amplitudes of SST.

@@

Mg/Ca elemental imaging was carried out using LIBS at the Leibniz Zentrum für Archäologie (Mainz-Germany) following previously published methods [@Hausmann2023-ih]. The imaging involves the ablation of the shell section using an infrared (1064 nm) laser (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume. This plume emits light, which is measured using a synchronised spectrometer (delay time = 0.5 µs, integration time = 1 µs). The resulting light spectrum quantifies the emission lines of Magnesium (MgII; 279.553 nm) and Calcium (CaII; 315.887 nm) to determine their intensity ratio. While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g. [@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca variation within the shell carbonate. All values reported here are in arbitrary units based on the ratio of intensities of the peaks above.

@@

Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O values and Mg/Ca ratios. DTW is an algorithm that measures similarity between two proxy sequences, which may vary in sampling resolution or interval. By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences. This allows us to compare the proxy datasets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates. We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data. This can involve selecting appropriate distance measures and constraints to ensure meaningful alignment. We used the default settings for symmetric step patterns and euclidean distance measures (see code in [@Hausmann_Surge_Godino_2024]).

@@

The elemental imaging of *P. vulgata* shells consistently showed repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca intensity ratios indicating high temperature periods and low ratios indicating low temperature periods. As examples, @fig-Pat_LIBS shows the 2D-distribution of elemental ratios across the entire section of ORK-LT5 and the preserved anterior side of QG1-7188-1 (additional maps can be found in the supplementaries). The repeating patterns are found across the calcitic layers that are exterior to the myostracum (layers m+2 and m+3) as well as the layers that are interior (m-2)(see Fenger et al. [-@Fenger2007-gf] for a detailed explanations of microstructural layers in *P. vulgata*). The layer m-1, which is also interior, consists of aragonite and is seen in @fig-Pat_LIBS in grey, as its Mg/Ca intensity ratio consistently falls below the range of interest (0.2 and above). Compared to other patelloid shells, some of whose interior is almost entirely made of aragonite, this layer is very thin. \newline

@@

Both specimens in @fig-Pat_LIBS show increased Mg/Ca intensity ratios towards the outside of the shell and some degree of intra-increment variability. This is particularly visible in @fig-Pat_LIBS b, where Mg/Ca intensity ratios range from \~0.5–0.7 in the first year of growth (9–12 mm on the y-axis). A similar pattern is visible in the interior m-2 layer of ORK-LT5 (@fig-Pat_LIBS a), which increase towards the anterior of the shell (0–10 mm on the x-axis).

@@

@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O values (similar graphs for other specimen can be found in the supplementaries). The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have a range of distances to the edge, depending on where one measures, and a line scan along the interior would be shorter than a line scan along the exterior. That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.

@@

Aligning the records programmatically using dynamic time warping, allowed us to compare both proxies for each specimen directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr). The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters ($\delta$^18^Ο = 0.6 – 0.065 \* Mg/Ca). The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2). These specimens are also the oldest and youngest specimens, respectively, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies. The mean R^2^ value of the individual *P. vulgata* specimens is 0.89, suggesting a generally good fit between the two proxies. P-values are overwhelmingly significant and consistently below p<1e^-6^.

```{r Patella_Correlation_Graphs,dev='png',dpi=300}
#| fig-cap: Correlation graphs for *Patella vulgata* specimens.
#| fig-height: 8
#| fig-width: 16
#| label: fig-Pat_Corr

@@

@@

```{r Nacella_LIBS,fig.pos='H'}
#| fig-cap: Example LIBS maps and line scans of analysed *Nacella* spp. specimens. (*a+c*) ND-1016-3 and (*b*+*d*) NM-1016-3.  Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast. Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range. Note the inset in (*c*) where the shell edge has been resampled using a higher resolution (50 µm instead of 100 µm) to better understand the Mg/Ca ratios at the shell edge. White dashed lines indicate location of line scans.
#| fig-height: 7
#| fig-width: 16
#| label: fig-Nac_LIBS

@@

Interestingly, the Mg/Ca intensity ratios in the *Nacella* spp. shells were much lower (0.05–0.30) than those seen in other patelloid species (e.g. 0.3–0.7 in *P. vulgata* above, or 0.5–1.5 in Hausmann *et al*. [-@Hausmann2023-ih]) using the same emission lines. While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen Magnesium to the chosen Calcium peak, the low ratios still indicate lower concentrations than seen elsewhere or seen only in aragonitic parts of e.g. *Patella* shells. Similar observations have been made before [@Graniero2017-io]. Strong intra-increment heterogeneities are visible along the exterior of the shells (m+3 layer), where Mg/Ca intensity ratios are higher, particularly in specimen ND-1016-3 (@fig-Nac_LIBS). The step from lower to higher along the increment is not gradual, as is the case in NM-1016-3 (see inset of @fig-Nac_LIBS b), but rather stepwise. Sampling across this boundary in a linear scan would most likely lead to erroneous results. Also sampling across gradual changes as in NM-1016-3 can influence the overall correlation graph for a specimen.

@@

```{r Nacella_Correlation_Graphs,dev='png',dpi=300}
#| fig-cap: Correlation graphs for *Nacella* spp. specimens
#| fig-height: 5
#| fig-width: 16
#| label: fig-Nac_Corr

@@

P_individual <- P_all+ylim(0,0.4)+
  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.35,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+
  theme_cowplot(20)+
    coord_fixed(ratio = 6) +
  geom_text(aes(x = 0.2, y = 0.3, label = "p-value < 0.0001", col = sampleID),size=6,hjust=0) +

>>>>

---
title: "Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Ivan Briz i Godino
    affiliations:
      - name: CONICET-CADIC
        address: AV. B. Houssay, 200.
        city: Ushuaia 
        country: Argentina
        postal-code: 9410 
abstract: |
  This study provides a reassessment of magnesium to calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST). While *Patella depressa* along the Spanish shoreline and *Patella caerulea* in the Mediterranean have repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species. Particularly, *Patella vulgata* but also *Nacella deaurata* have been studied using Mg/Ca ratios with mixed or contrary results. In this study, we present elemental maps of these two species as well as *Nacella magellanica* together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST. Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios. By reassessing these species and including modern and archaeological specimens (n=12) from three patelloid species (*P. vulgata*, *N. deaurata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change. As a result, these species can in the future serve to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for targeted $\delta$^18^O and clumped isotope analysis.
keywords: 
  - Sclerochronology
  - Patella
  - Nacella
  - LIBS
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: TRUE
    latex_engine: xelatex
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

@@

Limpet shells are commonly found within archaeological sites and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source [@Alvarez2011-rm, @Robson2023-oz]. They have been successfully studied in the context of coastal subsistence economies and site occupation on seasonal [@Shackleton1973-ij; @Parker2020-oy; @Bosch2018-ud] or long-term scales [@Ortiz2015-mr], as well as archives of palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl]. Determining past sea surface temperature (SST) change recorded in patelloid limpet shells largely relies on the measurement of $\delta$^18^O values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) ratios [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure that potentially provides a more accurate SST estimate than $\delta$^18^O-values,as the latter proxy is affected by changes in the mixing of fresh- and saltwater. In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].

@@

Recent research particularly of *Patella* spp. in the Mediterranean and southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf]. However, there remains a lack of clarity for Atlantic limpet species, particularly since it was shown, that the Mg/Ca ratio in their shells are not reliable recorders of palaeotemperature [@Graniero2017-io]. Using modern reference material Graniero *et al*. [-@Graniero2017-io] tested multiple elemental ratios (Mg/Ca, Sr/Ca, Li/Ca, Li/Mg, and Sr/Li) acquired via LA-ICP-MS measurements with no ratios showing correlations with corresponding SST changes nor any sinusoidal patterns that would indicate a seasonal environmental control of elemental ratios. This study included two species also used in our study (*Patella vulgata*, *Nacella deaurata*), with one *P. vulgata* specimen (i.e. specimen ORK-LT5) being analysed by both. We also added specimens of the species *Nacella magellanica,* which — similar to the other two species — is also an established palaeothermometer using $\delta$^18^O-values [@Colonese2012-ct; @gutiérrez-zugasti2017; @Nicastro2020-ih].

In this study, we repeat and expand upon the analysis of Atlantic limpets to determine the reliability of Mg/Ca ratios as palaeotemperature proxies. To do this, we sampled a set of previously published and unpublished limpet specimens dating to modern and archaeological contexts using LIBS. LIBS allows us to carry out 2D imaging of entire shell sections, which facilitates navigation of the complex elemental structure of the shell and a better separation of the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio. By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest. Generally establishing the usefulness and reliability of Mg/Ca ratios as SST proxy in limpets provides a platform for future research and an important steppingstone to a better understanding of elemental ratios in other marine mollusc shells.

@@

| Context        | Study              | Species          | Location        | Sample ID  | Previous analyses                     |
|----------|-----------------|------------|------------|---------|------------|
| Modern         | [@Nicastro2020-ih] | *N. deaurata*    | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C |
|                |                    |                  | \- Tierra del   | ND-1016-4  | analysis                              |
|                |                    | *N. magellanica* | Fuego (AR)      | NM-1016-1  |                                       |
|                |                    |                  |                 | NM-1016-3  |                                       |
|                |                    |                  |                 |            |                                       |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | $\delta$^18^O and $\delta$^13^C |
|                |                    |                  | Westray (UK)    |            | analysis;                             |
|                |                    |                  |                 |            | Mg, Li, Sr, Ca                       |
|                |                    |                  |                 |            |                                       |
|                |                    |                  |                 |            |                                       |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | $\delta$^18^O and $\delta$^13^C |
|                |                    |                  | Westray (UK)    | QG1-7189-2 | analysis                              |
|                |                    |                  |                 | QG2-1061-1 |                                       |
|                |                    |                  |                 | QG2-1064-1 |                                       |
|                |                    |                  |                 | QG2-7180-1 |                                       |
|                |                    |                  |                 | QG2-7180-2 |                                       |
|                |                    |                  |                 | QG1-7246-1 |                                       |

@@

Specimens were sourced from two main locations: the Beagle Channel in Tierra del Fuego (Argentina) and the island of Westray in the Orkney archipelago of Scotland (UK). Four modern specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016. The area receives around 570 mm of annual precipitation, but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu (practical salinity units) [@Nicastro2020-ih]. SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February (@fig-SSTs -- red line).

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for $\delta$^18^O and $\delta$^13^C values, as well as for elemental ratios via LA-ICP-MS [@Graniero2017-io]. Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho]. The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs -- blue line).
```{r SST_plot,fig.pos='H'}
#| fig-cap: Monthly sea surface temperature values averaged from 2007 to 2023. Values were sourced from @Good2020-nl.
#| fig-height: 2.5
#| fig-width: 6
#| label: fig-SSTs

@@

```
We also studied 7 archaeological specimens of *P. vulgata* from the Quoygrew site on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly (810 -- 1229 CE) [@Surge2012-ba].

@@

High-quality $\delta$^18^O values were acquired from previous publications on *P. vulgata* [@Surge2012-ba; @Graniero2017-io], *N. deaurata* and *N. magellanica* [@Nicastro2020-ih], in which detailed descriptions of the methodology for the stable isotope analyses can be found. In short, all shells were sectioned along the maximum growth axis to expose the internal growth structure. From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic layers of the shells (m+2 and m+3, see [@Fenger2007-gf]). The exact location of the sampled carbonate powder is not available apart from the distance to the growth edge, where reported. Analytical precision of the isotope ratios was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.

@@

@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species (*Nacella* spp., and *Patella vulgata*, respectively). Due to the high-resolution sampling, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* spp. shells living 2–3 years and *P. vulgata* living regularly over 3 years. In those instances where longer lifespans resulted in slow growth rates below 1 mm/yr (i.e. specimen QG1-7246-1, @fig-Pat_iso b), the temporal-resolution is reduced from more than a dozen of samples per year, to only around 7. Together they show a range of lifespans, with some comparatively long lived for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to within-shell time-averaging because of slow growth rates. That said, the majority yielded sufficiently fast growth to nearly capture seasonal amplitudes of SST with ages of 2--3 years (n=3), 3--4 years (n=2), and 5 years (n=2).

@@

Mg/Ca elemental imaging was carried out using LIBS at the Leibniz Zentrum für Archäologie (Mainz-Germany) following previously published methods [@Hausmann2023-ih]. The imaging involves the ablation of the shell section using an infrared (1064 nm) laser (1–-2 mJ, 100 Hz) onto a surface area of 20–-30 µm to generate a plasma plume. This plume emits light, which is measured using a synchronised spectrometer (delay time = 0.5 µs, integration time = 1 µs). The resulting light spectrum quantifies the emission lines of magnesium (MgII: 279.553 nm) and calcium (CaII: 315.887 nm) to determine their intensity ratio. While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g. [@Martinez-Minchero2023-pq]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca variation within the shell carbonate. All values reported here are in arbitrary units based on the ratio of intensities of the peaks above.

@@

Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O values and Mg/Ca ratios. DTW is an algorithm that measures similarity between two proxy sequences, which may vary in sampling resolution or interval. By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences. This allows us to compare the proxy datasets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates. We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data. This can involve selecting appropriate distance measures and constraints to ensure meaningful alignment. We used the default settings for symmetric step patterns and euclidean distance measures (see code in Hausmann et al.[-@hausmann2024confirmation]).

@@

The elemental imaging of *P. vulgata* shells consistently showed repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca intensity ratios indicating high temperature periods and low ratios indicating low temperature periods. As examples, @fig-Pat_LIBS shows the 2D-distribution of elemental ratios across the entire section of ORK-LT5 and the preserved anterior side of QG1-7188-1 (additional maps can be found in the supplementaries). The repeating patterns are found across the calcitic layers that are exterior to the myostracum (layers m+2 and m+3) as well as the layers that are interior (m--2)(see Fenger et al. [-@Fenger2007-gf] for a detailed explanations of microstructural layers in *P. vulgata*). The layer m-1, which is also interior, consists of aragonite and is seen in @fig-Pat_LIBS in grey, as its Mg/Ca intensity ratio consistently falls below the range of interest (0.2 and above). Compared to other patelloid shells, some of whose interior is almost entirely made of aragonite, this layer is very thin. \newline

@@

Both specimens in @fig-Pat_LIBS show increased Mg/Ca intensity ratios towards the outside of the shell and some degree of intra-increment variability. This is particularly visible in @fig-Pat_LIBS b, where Mg/Ca intensity ratios range from \~0.5--0.7 in the first year of growth (9-–12 mm on the y-axis). A similar pattern is visible in the interior m-2 layer of ORK-LT5 (@fig-Pat_LIBS a), which increase towards the anterior of the shell (0--10 mm on the x-axis).

@@

@@

```
@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O values (similar graphs for other specimen can be found in the supplementaries). The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have a range of distances to the edge, depending on where one measures, and a line scan along the interior would be shorter than a line scan along the exterior. That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.

```{r Patella_Correlation_Graphs,dev='png',dpi=300,fig.pos='H'}
#| fig-cap: Correlation graphs for Mg/Ca and $\delta$^18^O values in  *P. vulgata* specimens.
#| fig-height: 8
#| fig-width: 16
#| label: fig-Pat_Corr

@@

```
Aligning the records programmatically using dynamic time warping, allowed us to compare both proxies for each specimen directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr). The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters ($\delta$^18^Ο = 0.6 -- 0.065 \* Mg/Ca). The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2). These specimens are also the oldest and youngest specimens, respectively, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies. The mean R^2^ value of the individual *P. vulgata* specimens is 0.89, suggesting a generally good fit between the two proxies. P-values are overwhelmingly significant and consistently below p\<1e^--6^.

@@

```{r Nacella_LIBS,fig.pos='H'}
#| fig-cap: Example LIBS maps and line scans of analysed *Nacella* spp. specimens. (*a+c*) ND-1016-3 and (*b*+*d*) NM-1016-3.  Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast. Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range. Note the inset in (*b*) where the shell edge has been resampled using a higher resolution (50 µm instead of 100 µm) to better understand the Mg/Ca ratios at the shell edge. White dashed lines indicate location of line scans.
#| fig-height: 7
#| fig-width: 16
#| label: fig-Nac_LIBS

@@

Interestingly, the Mg/Ca intensity ratios in the *Nacella* spp. shells were much lower (0.05–0.30) than those seen in other patelloid species (e.g. 0.3–0.7 in *P. vulgata* above, or 0.5–1.5 in Hausmann *et al*. [-@Hausmann2023-ih]) using the same emission lines. While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen magnesium to the chosen calcium peak, the low ratios still indicate lower concentrations than seen elsewhere or seen only in aragonitic parts of e.g. *Patella* shells. Similar observations have been made before [@Graniero2017-io]. Strong intra-increment heterogeneities are visible along the exterior of the shells (m+3 layer), where Mg/Ca intensity ratios are higher, particularly in specimen ND-1016-3 (@fig-Nac_LIBS). The step from lower to higher along the increment is not gradual, as is the case in NM-1016-3 (see inset of @fig-Nac_LIBS b), but rather stepwise. Sampling across this boundary in a linear scan would most likely lead to erroneous results. Also sampling across gradual changes as in NM-1016-3 can influence the overall correlation graph for a specimen.

@@

```{r Nacella_Correlation_Graphs,dev='png',dpi=300}
#| fig-cap: Correlation graphs for Mg/Ca and $\delta$^18^O values in *Nacella* spp. specimens
#| fig-height: 5
#| fig-width: 16
#| label: fig-Nac_Corr

@@

P_individual <- P_all+ylim(0,0.4)+
  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.35,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+
  theme_cowplot(20)+
    coord_fixed(ratio = 6) +
  geom_text(aes(x = 0.2, y = 0.3, label = "p-value < 0.0001", col = sampleID),size=6,hjust=0,show.legend = FALSE) +
